Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-03-16 src/tex/ms.tex
[showyourwork] overleaf sync

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{1,}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{2,9}$,
    Sarah L. Casewell$^{3}$,
    Nicolas B. Cowan$^{4}$,
    David S. Doelman$^{5}$,\newauthor
    Timothy D. Gebhard$^{6}$,
    Matthew Kenworthy$^{5}$,
    Johan Mazoyer$^{7}$,
    Maxwell A. Millar-Blanchaer$^{8}$
    \newauthor [remaining workshop participants in alphabetical order] \newauthor 
    \\
    $^{1}$ Department of Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH, UK\\
    $^{2}$ Department of Earth and Planetary Sciences, University of California, Riverside, CA 92521, USA \\
    $^{3}$ Centre for Exoplanet Research, School of Physics and Astronomy, University of Leicester, University Road, Leicester, LE1 7RH, UK\\
    $^{4}$ Department of Earth \& Planetary Sciences and Department of Physics, McGill University, 3600 rue University, Montréal, QC, H3A 2T8, Canada\\
    $^{5}$ Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\\
    $^{6}$ Max Planck Institute for Intelligent Systems, Max Planck Ring 4, 72076 Tübingen, Germany \& ETH Zurich, Switzerland \\
    $^{7}$ LESIA, Observatoire de Paris, Université PSL, CNRS, Sorbonne Université, Université de Paris, F-92195 Meudon, France \\
    $^{8}$ Department of Physics, University of California, Santa Barbara, CA 93106, USA \\
    $^{9}$ NASA Nexus for Exoplanet System Science, Virtual Planetary Laboratory Team, Seattle, WA 98195, USA
}
\date{Draft version: \today}

@@

Surface liquid water is closely related to planetary habitability: water is required for life as we know it, and life on exoplanets will most likely on be detectable if it is present at the surface.

To first order, the presence of surface liquid water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which an Earth-like planet could maintain surface liquid water \citep{kasting93}.  The HZ is convenient because a planet can be deemed habitable based on solely on its orbital separation, but it remains an untested hypothesis. %Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

@@

\subsection{Ocean Glint}
\cite{2001Natur.412..885F} noted the potential for time-resolved photometry to identify water oceans on the surface of an exoplanet, which was further explored by \citet{williams2004} and \citet{giados2006}. Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., methane lakes on Titan;][]{2010GeoRL..37.7104S}. \citet{lustig2019} explored mapping exoplanets from their ocean glint in polarized light.

In polarized light the glint appears at similar phase angles but its peak can be extended further particularly in the presence of atmospheric winds \citep{kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean.
\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.

>>>>

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{1,}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{2,28}$,
    Sarah L. Casewell$^{3}$,
    Nicolas B. Cowan$^{4}$,
    David S. Doelman$^{5}$,\newauthor 
    Timothy D. Gebhard$^{6}$,
    Matthew Kenworthy$^{5}$,
    Johan Mazoyer$^{7}$,
    Maxwell A. Millar-Blanchaer$^{8}$,\newauthor 
    Olivier Absil$^{16}$,
    Lisa Altinier$^{13}$,
    Pierre Baudoz$^{7}$,
    Ruslan Belikov$^{24}$,
    Alexis Bidot$^{9}$,\newauthor 
    Markus Johannes Bonse$^{6}$,
    Bernhard Brandl$^{5}$,
    Alexis Carlotti$^{9}$,
    Elodie Choquet$^{13}$,
    Niyati Desai$^{17}$,\newauthor
    Kevin Fogarty$^{24}$,
    Jules Fowler$^{12}$
    Yann Gutierrez$^{7,14,22}$,
    Olivier Guyon$^{15,25,26,27}$,
    Sebastiaan Y. Haffert,$^{15}$, \newauthor
    Olivier Herscovici-Schiller$^{14}$,
    Adrien Hours$^{9}$,
    Roser Juanola-Parramon $^{20,21}$,
    Elina Kleisioti $^{5}$, \newauthor
    Lorenzo König$^{16}$,
    Mariya Krasteva$^{23}$,
    Iva Laginja$^{7}$,
    Rico Landman$^{5}$,
    Lucie Leboulleux$^{9}$,
    David Mouillet$^{9}$,\newauthor
    Mamadou N’Diaye$^{10}$,
    Emiel H. Por$^{18}$,
    Laurent Pueyo$^{18}$,
    Frans Snik$^{5}$,
    Dirk van Dam$^{5}$,
    Kyle van Gorkom, $^{15}$,\newauthor
    Maaike van Kooten$^{11}$
    \newauthor 
    \\
    $^{1}$ Department of Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH, UK\\
    $^{2}$ Department of Earth and Planetary Sciences, University of California, Riverside, CA 92521, USA \\
    $^{3}$ Centre for Exoplanet Research, School of Physics and Astronomy, University of Leicester, University Road, Leicester, LE1 7RH, UK\\
    $^{4}$ Department of Earth \& Planetary Sciences and Department of Physics, McGill University, 3600 rue University, Montréal, QC, H3A 2T8, Canada\\
    $^{5}$ Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\\
    $^{6}$ Max Planck Institute for Intelligent Systems, Max Planck Ring 4, 72076 Tübingen, Germany \& ETH Zurich, Switzerland \\
    $^{7}$ LESIA, Observatoire de Paris, Universit\'e PSL, CNRS, Sorbonne Universit\'e, Universit\'e de Paris, F-92195 Meudon, France \\
    $^{8}$ Department of Physics, University of California, Santa Barbara, CA 93106, USA \\
$^{9}$ Univ. Grenoble Alpes, CNRS, IPAG, 38000 Grenoble, France \\
$^{10}$ Universit\'e C\^ote d'Azur, Observatoire de la C\^ote d'Azur, CNRS, Laboratoire Lagrange. France \\
$^{11}$ National Research Council Canada, Herzberg Astronomy and Astrophysics Research Center, Victoria, B.C. Canada \\
$^{12}$ University of California, Santa Cruz  \\
$^{13}$ Aix Marseille Univ, CNRS, CNES, LAM, Marseille, France \\
$^{14}$ DTIS, ONERA, Université Paris Saclay, 91123 Palaiseau, France \\
$^{15}$ Steward Observatory, University of Arizona, 933 North Cherry Avenue, Tucson, Arizona \\
$^{16}$ STAR Institute, Universit\'e de Li\`ege, All\'ee du Six Ao\^ut 19c, 4000 Li\`ege, Belgium \\
$^{17}$ Department of Astronomy, California Institute of Technology, Pasadena, CA, 91125, USA \\
$^{18}$ Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA \\
$^{19}$ SRON Netherlands Institute for Space Research, Niels Bohrweg 4, 2333 CA, Leiden, The Netherlands \\
$^{20}$ NASA Goddard Space Flight Center, 8800 Greenbelt Rd, Greenbelt, MD  20771,United States \\
$^{21}$ University of Maryland Baltimore County, 1000 Hilltop Cir, Baltimore, MD 21250, United States \\
$^{22}$ DOTA, ONERA, 92322 Ch\^atillon, France \\
$^{23}$ European Space Agency, ESTEC, Keplerlaan 1, 2200 AG Noordwijk, The Netherlands \\
$^{24}$ NASA Ames Research Center, Bldg. 245, Moffett Field, USA \\
$^{25}$ Subaru Telescope, NAOJ, United States \\
$^{26}$ College of Optical Sciences, University of Arizona, Tucson, AZ 87521, United States \\
$^{27}$ Astrobiology Center, 2 Chome-21-1, Osawa, Mitaka, Tokyo, 181-8588, Japan \\
$^{28}$ NASA Nexus for Exoplanet System Science, Virtual Planetary Laboratory Team, Seattle, WA 98195, USA}
\date{Draft version: \today}

@@

Surface liquid water is closely related to planetary habitability: water is required for life as we know it, and life on exoplanets will most likely on be detectable if it is present at the surface. 
To first order, the presence of surface liquid water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which an Earth-like planet could maintain surface liquid water \citep{kasting93}.  The HZ is convenient because a planet can be deemed habitable based on solely on its orbital separation, but it remains an untested hypothesis. We would therefore like to establish the presence of liquid water on the surface of an exoplanet before trying to interpret its atmospheric biosignatures. %Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

@@

\subsection{Ocean Glint}
\cite{2001Natur.412..885F} noted the potential for time-resolved reflected light photometry to identify water oceans on the surface of an exoplanet.  \cite{2008Icar..195..927W} performed simulations of scattered light from directly imaged terrestrial exoplanets. 
%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. %Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but
Specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., methane lakes on Titan;][]{2010GeoRL..37.7104S}. %\citet{lustig2019} explored mapping exoplanets at crescent phases using ocean glint.

In polarized light the glint appears at similar phase angles but its peak can be extended further particularly in the presence of atmospheric winds \citep{kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean.
%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.
